Software for the analysis and visualization of deep mutational scanning data

### 2015-03-23 paper/BMC_Bioinformatics_2015/ResponseToReviewers.tex

Bloom describes a software package for analysis of deep mutational scanning using a novel (for this application) Bayesian/MCMC framework. The general framework appears to be sound and addresses a significant need for these new and highly promising functional assays. The manuscript is well-written and explains the major uses cases of the software in sufficient detail.

Minor comments/suggestions:

1. The notion that differences based on larger absolute counts provide more certainty than those based on small absolute counts is obviously true, but only up to a point. Since deep mutational scanning experiments typically involves exponential amplification of a finite number of selected molecules prior to sequencing, there will be a saturation point (or region) beyond which higher absolute counts no longer provide additional information. This effect might not be negligible in experiments that involve typical cultures of $10^6$ to $10^8$ independently transformed or transduced cells. I would encourage the author to note these effects and perhaps add features or provide guidance to software users for how to estimate or detect the saturation point.

\begin{response}
This is an excellent point. The text has been revised to emphasize that increasing deep-sequencing counts stop being meaningful once the typical number of counts per site becomes on the order of the number of unique molecules that were originally in the experiment. For most shotgun sequencing experiments, the typical counts per site ($\sim 10^5$ to $10^6$) still tend to be less than the typical number of unique molecules ($\sim 10^6$ to $10^8$), but this may not always be true. The revised text suggests that the experiments should be done in a way that ensures that the sequencing depth is not saturating the number of unique molecules (i.e. by quantifying the number of molecules carefully). It suggests that one way that this issue can be examined during \textit{post hoc} computational analyses is by randomly subsampling the deep sequencing data (when generating the mutation counts files used as input to inference programs in \textsl{dms\_tools}), and then seeing if this reduced depth alters the results. 
\end{response}

@@

2. On pg 2 – the notion of a mutation being ``beneficial'' is generally taken to be relative to the fitness of the wild-type (non-mutated) gene. Eq (1) is such that $\phi_{r,x} > 1$ does not imply that $x$ is beneficial in this sense. Consider the trivial example of a two site gene with two characters. The input library has one member with a neutral mutation at site 1, one member with a deleterious mutation at site 2 and one WT member. Post-selection $\phi$ for the non-WT character at site 1 will be $\frac{1/1}{1/2} > 1$ even though the non-WT character at site 1 is not beneficial. This should be clarified.

\begin{response}
This is a \emph{hugely} important point! The first paragraph of the subsection in the \textsl{Background} entitled \textsl{The goal: inferring site-specific amino-acid preferences} now discusses this issue in detail. Essentially, assuming the length $L$ of the gene is $\gg 1$ (which is the case for any reasonable gene), then if the mutant library contains a Poisson distribution of the number of mutations per gene (as is often the case), the problem noted by the reviewer will not arise in practice. The reason is that for such genes, knowing that there is a mutation at site $r$ does not appreciably change the marginal distribution of the number of mutations at other sites (it only changes it by a factor of $\frac{L - 1}{L}$, which in general will be very small). On the other hand, if the mutant library is created such that it contains genes with exactly one mutation per gene (which is now sometimes done), then the problem noted by the reviewer will arise even for realistic genes with $L \gg 1$. The subsection \textsl{The goal: inferring site-specific amino-acid preferences} discusses this issue, and how the problem can be corrected by appropriately defining the frequencies for the wildtype character.
\end{response}

@@

Question 1:
On page 2, in the second paragraph, a deep mutational scanning study is described. The average mutation rate is $\mu=1/L$, where $L$ is the length of the gene. It is unclear to me why this is so -- intuitively one might think that a longer gene may have more mutations, given that mutation is a random process. More generally, could the author provide more details on the range of data one would expect from typical deep mutational scanning experiments?

\begin{response}
In principle, a deep mutational scanning experiment could introduce any number of mutations per gene. But typically, experimentalists tend to introduce an average of $\approx$1 mutation per gene when they create the codon-mutant libraries. The reason is that having genes with $\gg$1 mutations makes it hard to attribute the selection on any given gene variant to any specific mutation. For this reason, published deep mutational scanning experiments have generally targeted $\overline{\mu} \sim 1/L$. This is an important distinction between deep mutational scanning and naturally occurring mutations -- naturally occurring mutations will be more common in longer genes, since typically the per-site mutation rate is approximately uniform and longer genes have more sites. The revised text now clarifies this point.
\end{response}

Question 2:
The author states that for a depth of $10^6$, we expect about 2000 counts of non-wild type codons for the average gene. The author arrived at this number by multiplying the total number of reads by the probability of a mutation. This is reasonable if each read comes from a different DNA fragment, each with probability $\mu$ of mutating. However, if multiple reads all cover the same DNA fragment, would this relationship still hold? The author should comment on whether this situation is likely to occur in practice and, if so, how it affects the method.

\begin{response}
This is an excellent comment, and is closely related to point (1) made by the first reviewer. The calculation of the counts to which the reviewer refers, and indeed much of the statistical framework described in the manuscript, is made under the assumption that the number of unique DNA fragments in the library substantially exceeds the number of sequencing reads, such that the deep-sequencing reads can be viewed as subsampling a very large number of DNA fragments. This issue has been clarified in the text as described in the response to point (1) made by the first reviewer.
\end{response}

### 2015-03-23 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article.tex
Responded to reviewer 1 comment 2

\section*{Background}
Deep mutational scanning is a high-throughput experimental technique to assess the impacts of mutations on a protein-coding gene~\cite{fowler2014deep}. Figure \ref{fig:DeepMutationalScan} shows a schematic of deep mutational scanning. A gene is mutagenized, and the library of resulting variants is introduced into cells or viruses, which are then subjected to an experimental selection that enriches for functional variants and depletes non-functional ones. Deep sequencing of the variants pre- and post-selection provides information about the functional impacts of mutations. Since the original description of deep mutational scanning by Fowler \textit{et al}~\cite{fowler2010high}, the technique has been applied to a wide range of genes~\cite{traxlmayr2012construction,mclaughlin2012spatial,starita2013activity,melamed2013deep,roscoe2013analyses,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,thyagarajan2014inherent,wu2014high,wu2014ns,olson2014comprehensive}, both to measure mutational tolerance given a single selection pressure as in Figure \ref{fig:DeepMutationalScan}A, or to identify mutations that have different effects under alternative selections as in Figure \ref{fig:DeepMutationalScan}B. New techniques to create comprehensive codon-mutant libraries of genes make it possible to profile all single amino-acid mutations~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive}, while new techniques for targeted mutagenesis of mammalian genomes enable deep mutational scanning to be applied across the biological spectrum from viruses and bacteria to human cells~\cite{findlay2014saturation}.

@@

\subsection*{The goal: inferring site-specific amino-acid preferences.}
The goal is to estimate the effects of mutations from changes in their counts after selection. Let $\mu_{r,x}$, $f_{r,x}$, $f_{r,x}^{s1}$, and $f_{r,x}^{s2}$ denote the \emph{true} frequencies of $x$ at $r$ that would be observed for the four libraries in Figure \ref{fig:DeepMutationalScan} if we sampled at infinite depth in both the actual experiment and the sequencing. If we knew these frequencies, we could calculate parameters that reflect the effects of mutations. One parameter that characterizes the effect of mutating $r$ from $\operatorname{wt}\left(r\right)$ to $x$ for the experiment in Figure \ref{fig:DeepMutationalScan}A is the \emph{enrichment ratio}, which is the relative frequency of mutations to $x$ after selection versus before selection:
\begin{equation}
\label{eq:enrichment}
\phi_{r,x} = \frac{f_{r,x} / f_{r,\operatorname{wt}\left(r\right)}}{\mu_{r,x} / \mu_{r,\operatorname{wt}\left(r\right)}}.
\end{equation}
Beneficial mutations have $\phi_{r,x} > 1$, while deleterious ones have $\phi_{r,x} < 1$. A related parameter is the \emph{preference} $\pi_{r,x}$ of $r$ for $x$. At each site, the preferences are simply the enrichment ratios rescaled to sum to one:
\begin{equation}
\label{eq:preference}
\pi_{r,x} = \frac{\phi_{r,x}}{\sum_y \phi_{r,y}} = \frac{f_{r,x} / \mu_{r,x}}{\sum_y f_{r,y} / \mu_{r,y}},
\end{equation}
or equivalently
\begin{equation}
\label{eq:preference2}
f_{r,x} = \frac{\pi_{r,x} \times \mu_{r,x}}{\sum_y \pi_{r,y} \times \mu_{r,y}},
\end{equation}
where $y$ is summed over all character identities (all nucleotides, codons, or amino acids).
The preferences can be intuitively visualized (Figure \ref{fig:DeepMutationalScan}A) and interpreted as the equilibrium frequencies in substitution models for gene evolution~\cite{bloom2014experimentally,bloom2014informed} (after accounting for uneven mutational rates~\cite{yampolsky2005exchangeability,stoltzfus2009climbing}).

@@

Our prior assumption is that the mutagenesis introduces all mutant characters at equal frequency (this assumption is only plausible for codons if the mutagenesis is at the codon level as in~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive}; if mutations are made at the nucleotide level such as by error-prone PCR then characters should be defined as nucleotides). The average per-site mutagenesis rate is estimated as
\begin{equation}
\label{eq:avgmu}
\overline{\mu} = \left(\frac{1}{L}\sum\limits_r \frac{1}{N_r^{\textrm{pre}}}\sum\limits_{x\ne \operatorname{wt}\left(r\right)} n_{r,x}^{\textrm{pre}}\right) - \sum\limits_{m \ge 1} \overline{\epsilon_m},
\end{equation}
so that
\begin{equation}
\boldsymbol{\mathbf{a_{r,\mu}}} = \left(\cdots, \frac{\overline{\mu}}{\mathcal{N}_x - 1} + \delta_{x,\operatorname{wt}\left(r\right)} \times \left[1 - \overline{\mu}\right] ,\cdots\right). \label{eq:armu}
\end{equation}

>>>>

\section*{Background}
Deep mutational scanning is a high-throughput experimental technique to assess the impacts of mutations on a protein-coding gene~\cite{fowler2014deep}. Figure \ref{fig:DeepMutationalScan} shows a schematic of deep mutational scanning. A gene is mutagenized, and the library of resulting variants is introduced into cells or viruses, which are then subjected to an experimental selection that enriches for functional variants and depletes non-functional ones. Deep sequencing of the variants pre- and post-selection provides information about the functional impacts of mutations. Since the original description of deep mutational scanning by Fowler \textit{et al}~\cite{fowler2010high}, the technique has been applied to a wide range of genes~\cite{traxlmayr2012construction,mclaughlin2012spatial,starita2013activity,melamed2013deep,roscoe2013analyses,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,thyagarajan2014inherent,wu2014high,wu2014ns,olson2014comprehensive,kitzman2015massively}, both to measure mutational tolerance given a single selection pressure as in Figure \ref{fig:DeepMutationalScan}A, or to identify mutations that have different effects under alternative selections as in Figure \ref{fig:DeepMutationalScan}B. New techniques to create comprehensive codon-mutant libraries of genes make it possible to profile all single amino-acid mutations~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,kitzman2015massively}, while new techniques for targeted mutagenesis of mammalian genomes enable deep mutational scanning to be applied across the biological spectrum from viruses and bacteria to human cells~\cite{findlay2014saturation}.

@@

\subsection*{The goal: inferring site-specific amino-acid preferences.}
The goal is to estimate the effects of mutations from changes in their counts after selection. Let $\mu_{r,x}$, $f_{r,x}$, $f_{r,x}^{s1}$, and $f_{r,x}^{s2}$ denote the \emph{true} frequencies at site $r$ of all mutant characters $x \ne \operatorname{wt}\left(r\right)$ that would be observed for the four libraries in Figure \ref{fig:DeepMutationalScan} if we sampled at infinite depth in both the actual experiment and the sequencing. The definition of these frequencies for the wildtype character $\operatorname{wt}\left(r\right)$ depends on how the mutant library is constructed. If the mutant library is constructed so that there is a Poisson distribution of the number of mutations per gene (as is the case for error-prone PCR or the codon-mutagenesis in \cite{bloom2014experimentally,thyagarajan2014inherent}), then $\mu_{r,\operatorname{wt}\left(r\right)}$, $f_{r,\operatorname{wt}\left(r\right)}$, $f_{r,\operatorname{wt}\left(r\right)}^{s1}$, and $f_{r,\operatorname{wt}\left(r\right)}^{s2}$ are defined as for all other characters $x$, and denote the frequencies of $\operatorname{wt}\left(r\right)$ at site $r$ that would be observed if sampling at infinite depth. The reason we can make this definition for libraries containing genes with Poisson-distributed numbers of mutations is that for any reasonable-length gene ($L \gg 1$), the marginal distribution of the number of mutations in a gene is virtually unchanged by the knowledge that there is a mutation at site $r$. On the other hand, if the mutant library is constructed so that there is exactly zero or one mutation per gene (as in \cite{firnberg2014comprehensive,melnikov2014comprehensive,kitzman2015massively}), then the marginal distribution of the total number of mutations in a gene is changed by the knowledge that there is a mutation at $r$. In this case, the wildtype-character frequencies $\mu_{r,\operatorname{wt}\left(r\right)}$, $f_{r,\operatorname{wt}\left(r\right)}$, $f_{r,\operatorname{wt}\left(r\right)}^{s1}$, and $f_{r,\operatorname{wt}\left(r\right)}^{s2}$ are correctly defined as the frequency of unmutated genes in the library, and the counts $n_{r,\operatorname{wt}\left(r\right)}^{\rm{pre}}$, etc. are defined as the number of reads at $r$ attributable to unmutated genes. In this case, measurement of these counts requires sequencing with linkage as in \cite{hiatt2010parallel,wu2014hiv,kitzman2015massively}. The proper analysis of libraries containing only unmutated and singly mutated clones sequenced without linkage is beyond the scope of this paper.

If we knew the frequencies $\mu_{r,x}$, $f_{r,x}$, $f_{r,x}^{s1}$, and $f_{r,x}^{s2}$, we could calculate parameters that reflect the effects of mutations. One parameter that characterizes the effect of mutating $r$ from $\operatorname{wt}\left(r\right)$ to $x$ for the experiment in Figure \ref{fig:DeepMutationalScan}A is the \emph{enrichment ratio}, which is the relative frequency of mutations to $x$ after selection versus before selection:
\begin{equation}
\label{eq:enrichment}
\phi_{r,x} = \frac{f_{r,x} / f_{r,\operatorname{wt}\left(r\right)}}{\mu_{r,x} / \mu_{r,\operatorname{wt}\left(r\right)}}.
\end{equation}
Beneficial mutations have $\phi_{r,x} > 1$, while deleterious ones have $\phi_{r,x} < 1$. A related parameter is the \emph{preference} $\pi_{r,x}$ of $r$ for $x$. At each site, the preferences are simply the enrichment ratios rescaled to sum to one:
\begin{equation}
\label{eq:preference}
\pi_{r,x} = \frac{\phi_{r,x}}{\sum_y \phi_{r,y}} = \frac{f_{r,x} / \mu_{r,x}}{\sum_y f_{r,y} / \mu_{r,y}},
\end{equation}
or equivalently
\begin{equation}
\label{eq:preference2}
f_{r,x} = \frac{\pi_{r,x} \times \mu_{r,x}}{\sum_y \pi_{r,y} \times \mu_{r,y}},
\end{equation}
where $y$ is summed over all character identities (all nucleotides, codons, or amino acids).
The preferences can be intuitively visualized (Figure \ref{fig:DeepMutationalScan}A) and interpreted as the equilibrium frequencies in substitution models for gene evolution~\cite{bloom2014experimentally,bloom2014informed} (after accounting for uneven mutational rates~\cite{yampolsky2005exchangeability,stoltzfus2009climbing}).

@@

Our prior assumption is that the mutagenesis introduces all mutant characters at equal frequency (this assumption is only plausible for codons if the mutagenesis is at the codon level as in~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,kitzman2015massively}; if mutations are made at the nucleotide level such as by error-prone PCR then characters should be defined as nucleotides). The average per-site mutagenesis rate is estimated as
\begin{equation}
\label{eq:avgmu}
\overline{\mu} = \left(\frac{1}{L}\sum\limits_r \frac{1}{N_r^{\textrm{pre}}}\sum\limits_{x\ne \operatorname{wt}\left(r\right)} n_{r,x}^{\textrm{pre}}\right) - \sum\limits_{m \ge 1} \overline{\epsilon_m},
\end{equation}
so that
\begin{equation}
\boldsymbol{\mathbf{a_{r,\mu}}} = \left(\cdots, \frac{\overline{\mu}}{\mathcal{N}_x - 1} + \delta_{x,\operatorname{wt}\left(r\right)} \times \left[1 - \overline{\mu}\right] ,\cdots\right). \label{eq:armu}
\end{equation}
